Homework 5 (due 22 Oct in class)
========================================================

### Find the nugget, sill, and range (or effective range) for the linear, spherical, exponential, and Gaussian covariance function. ###

First consider the __linear__ case:

$$
\gamma(t) = \begin{cases} 
      \tau^2 + \sigma^2 t  & \text{if } t >0, \sigma >0, \tau >0 \\
      0 & \text{otherwise}
   \end{cases}
$$

* $\tau^2$ is the _nugget_ since $\lim_{t \rightarrow 0^+} \gamma(t) = \tau^2$.

* Note that the _sill_ infinite since $\lim_{t \rightarrow \infty} \gamma(t) = \infty$.

* Since the _sill_ is infinite, the _range_ is also infinite.

Next consider the __spherical__ case:

$$
\gamma(t) = \begin{cases} 
      \tau^2 + \sigma^2  & \text{if } t \geq 1/\phi \\
      \tau^2 + \sigma^2 (\frac{3\phi t}{2} - \frac{(\phi t)^3}{2})  & \text{if } 0 < t < 1/\phi \\
      0 & \text{otherwise}
   \end{cases}
$$

* $\tau^2$ is the _nugget_ since $\lim_{t \rightarrow 0^+} \gamma(t) = \tau^2$.

* $\tau^2 + \sigma^2$ is the _sill_ since $\lim_{t \rightarrow \infty} \gamma(t) = \tau^2 + \sigma^2$.

* $1/\phi$ is the _range_ since $\gamma(t)$ reaches the _sill_ when $t = 1/\phi$.

Next consider the __exponential__ case:

$$
\gamma(t) = \begin{cases} 
      \tau^2 + \sigma^2 (1 - exp\{ -\phi t \})  & \text{if } t > 0 \\
      0 & \text{otherwise}
   \end{cases}
$$

* $\tau^2$ is the _nugget_ since $\lim_{t \rightarrow 0^+} \gamma(t) = \tau^2$.

* $\tau^2 + \sigma^2$ is the _sill_ since $\lim_{t \rightarrow \infty} \gamma(t) = \tau^2 + \sigma^2$.

* Since the _sill_ is reached asymptotically, the _range_ must be defined in terms of _effective range_. That is, we have to find the value of $t$ such that the covariance is less than 0.05. Now, note that

$$
\begin{align*}
C(t) &= \lim_{u \rightarrow \infty} \gamma(u) - \gamma(t)
\\ &= \tau^2 + \sigma^2 - [\tau^2 + \sigma^2(1 - exp\{ - \phi t\})]
\\ &= \sigma^2exp\{ - \phi t\}
\end{align*}
$$

By setting $C(t) = 0.05$, we obtain the _effective range_ of $t = 3/\phi$

Next consider the __gaussian__ case:

$$
\gamma(t) = \begin{cases} 
      \tau^2 + \sigma^2 (1 - exp\{ -\phi^2 t^2 \})  & \text{if } t > 0 \\
      0 & \text{otherwise}
   \end{cases}
$$

* $\tau^2$ is the _nugget_ since $\lim_{t \rightarrow 0^+} \gamma(t) = \tau^2$.

* $\tau^2 + \sigma^2$ is the _sill_ since $\lim_{t \rightarrow \infty} \gamma(t) = \tau^2 + \sigma^2$.

* Since the _sill_ is reached asymptotically, the _range_ must be defined in terms of _effective range_. That is, we have to find the value of $t$ such that the covariance is less than 0.05. Now, note that

$$
\begin{align*}
C(t) &= \lim_{u \rightarrow \infty} \gamma(u) - \gamma(t)
\\ &= \tau^2 + \sigma^2 - [\tau^2 + \sigma^2(1 - exp\{ -\phi^2 t^2 \})]
\\ &= \sigma^2exp\{ -\phi^2 t^2 \}
\end{align*}
$$

By setting $C(t) = 0.05$, we obtain the _effective range_ of $t = \sqrt{3}/\phi$



### Plot the variograms for the linear, spherical, exponential, and Gaussian using the same value for the nugget, sill, and range (or effective range). ###

```{r setup, echo=FALSE, message=FALSE}
library(knitr)
render_html()
```

```{r plots, fig.width=12}
sigma2 <- .5
tau2 <- .2
t <- seq(from=0.001, to=2, by=.01)
linear <- tau2 + sigma2*t
phi <- 1
spherical <- ifelse(t >= 1/phi, tau2 + sigma2, tau2 + sigma2*(3*phi*t/2 - (.5)*(phi*t)^3))
phi_exp <- 3/phi
exponential <- tau2 + sigma2*(1 - exp(-phi_exp*t))
phi_gauss <- sqrt(3)/phi
gauss <- tau2 + sigma2*(1 - exp(-phi_gauss^2*t^2))
plot(t, linear, type="l", main="Theoretical semivariogram", ylim=c(0, 1), ylab=expression(gamma(t)))
lines(t, spherical, col=2)
lines(t, exponential, col=3)
lines(t, gauss, col=4)
legend(1.1, 0.6, c("linear", "spherical", "exponential", "gaussian"), col=1:4, lty=1)
```


### Use Brook's Lemma to find the joint distribution for the conditional distributions $X|Y \sim N(\rho Y, 1-\rho^2)$ and $Y|X \sim N(\rho X, 1-\rho^2)$. ###

According to Brook's Lemma, for any $X'$ and $Y'$, 

$$
\frac{p(X, Y)}{p(X', Y')} = \frac{p(X|Y)}{p(X'|Y)}\frac{p(Y|X')}{p(Y'|X')}
$$

Choosing $X', Y' = 0$ we have 

$$
\begin{align*}
\frac{p(X, Y)}{p(X', Y')} &\propto \frac{ exp\{ \frac{-(x - \rho y)^2 - y^2}{2(1-\rho^2)} \} }{ exp\{ \frac{-(\rho y)^2}{2(1-\rho^2)} \} }
\\ &\propto exp\{ \frac{ -x^2 + 2\rho xy - y^2}{2(1-\rho^2)} \}
\\ &= exp\{ - \frac{1}{2} (\begin{array}{cc} x&y \end{array}) \Sigma^{-1} (\begin{array}{c} x \\ y \end{array}) \}
\end{align*}
$$

where $\Sigma^{-1} = (\begin{array}{cc} \frac{1}{1 - \rho^2} & \frac{\rho}{1 - \rho^2} \\ \frac{\rho}{1 - \rho^2} & \frac{1}{1 - \rho^2}\end{array})$, which implies $\Sigma =  (\begin{array}{cc} 1 & \rho \\ \rho & 1 \end{array})$. Therefore, $(\begin{array}{cc} x \\ y \end{array}) \sim N(0, \Sigma)$