---
title: Performance and Salary in Baseball
author: "Carson Sievert"
date: "April 8, 2014"
output: html_document
---

Most would agree that work compensation should be based solely on performance. However, this isn't always the case. Is there an association between performance and compensation in baseball? Here we take the top 25 earners in terms of 2013 salary and look for any association between salary and performance amongst these players. Note that performance is measured by a statistic call Wins Above Replacement (WAR) player. In short, this measures how many games a player helped their team win.

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r getDat, cache=TRUE}
library(XML)
doc <- htmlParse("http://sports.newsday.com/long-island/data/baseball/mlb-salaries-2013/")
#get the actual salary numbers
nodes <- getNodeSet(doc, "//td[@class='sdb sdb-numeric sdb-field-2013salary']") 
sals <- sapply(nodes, xmlValue)
tmp <- gsub("\\$||,", "", sals)
salary <- as.numeric(tmp)/1e6 #in millions
nodes2 <- getNodeSet(doc, "//td[@class='sdb-last sdb-numeric sdb-field-age']") 
ages <- sapply(nodes2, xmlValue)
nodes3 <- getNodeSet(doc, "//td[@class='sdb sdb-field-player']")
names <- sapply(nodes3, xmlValue)
dat <- data.frame(names, ages, salary)

# csv files copy pasted from -- http://www.baseball-reference.com/leagues/AL/2013-value-batting.shtml")
stats_al <- cbind(read.csv("2013_AL.csv", header=TRUE, stringsAsFactors=FALSE), league="AL")
stats_nl <- cbind(read.csv("2013_NL.csv", header=TRUE, stringsAsFactors=FALSE), league="NL")
stats <- rbind(stats_al, stats_nl)
names(stats) <- gsub("\\.", "", names(stats))
stats$Salary <- as.numeric(gsub("\\$||,", "", stats$Salary))
o <- order(stats$Salary, decreasing=TRUE)
top_guys <- head(stats[o,], 25)
top_guys$Salary <- as.numeric(gsub("\\$||,", "", top_guys$Salary))
top_guys2 <- top_guys
top_guys2$Salary <- as.numeric(gsub("\\$||,", "", top_guys2$Salary))/1e6
```

```{r plot, eval=FALSE}
plot(top_guys$WAR, top_guys$Salary)
```

```{r chart, results='asis'}
library(rCharts)
r1 <- rPlot(Salary ~ WAR, data = top_guys, type = "point", tooltip="#! function(item){ return item.X } !#")
ymin <- 1.5e7
ymax <- 3e7
r1$addParams(height = 700, title = "Scatterplot of salary and performance amongst top 25 salary earners in MLB.")
r1$guides(x = list(min=-2, max=10, title = "Wins Above Replacement (WAR)"), 
          y = list(min=ymin, max=ymax, title="Salary (in dollars)"))
add <- data.frame(x = 0, y = c(ymin, max(top_guys$Salary)))
r1$layer(y ~ x,data = add, type="line")
r1$print('chart2', include_assets = TRUE)
```
