---
layout: post
title: "State of the pitchRx"
author: [cpsievert]
categories: [pitchRx, R]
tags: []
---
{% include JB/setup %}

Well, they've done it again. A few days ago, [@hadleywickham](https://twitter.com/hadleywickham/)
and [@rstudioapp](https://twitter.com/rstudioapp) provided another tool to help the layman look
like a hacker. This time, they've [provided
access](http://blog.rstudio.org/2013/06/10/rstudio-cran-mirror/) to logs of CRAN package downloads.
I've already noticed a [good](http://ramnathv.github.io/rstudio_logs/)
[amount](http://spatial.ly/2013/06/r_activity/)
[of](http://www.r-statistics.com/2013/06/answering-how-many-people-use-my-r-package/)
[analysis](http://www.r-statistics.com/2013/06/top-100-r-packages-for-2013-jan-may/) surrounding
this data. The announcement grabbed my attention because I released my first R package (`pitchRx`)
in Feburary and was curious about what type of activity it receives. Below is a plot of the number
of downloads (log transformed) for every package since Feburary 20th (the day `pitchRx` was
released). Note that the packages are ordered from lowest number of downloads to the highest. The
vertical line indicates the position of `pitchRx`.

```{r pitchRx, cache=TRUE, echo=FALSE}
setwd("~/Desktop/cran_logs")
df <- read.csv("2013.csv")
pitchRx <- df[as.character(df$package) == "pitchRx",]
pitchRx2 <- pitchRx[!is.na(pitchRx$package == "pitchRx"),]
```

```{r other, fig.width=11, cache=TRUE, echo=FALSE}
start <- min(as.Date(pitchRx2$date))
other <- df[as.Date(df$date) > start, ]
other$pitchRx <- as.numeric(other$package == "pitchRx")
tab.package <- table(other$package)
o <- order(tab.package)
other$package <- factor(other$package, levels=levels(other$package)[o])
packages <- names(tab.package)
tab.df <- data.frame(package=factor(packages, levels=packages[o]), Freq=as.numeric(tab.package))
ggplot(data=tab.df)+geom_point(aes(x=package, y=log(Freq)))+geom_vline(aes(xintercept=which(levels(package) %in% "pitchRx")))+theme(axis.text.x=element_blank(), axis.ticks.x=element_blank())+geom_text(label="pitchRx", aes(x=which(levels(package) %in% "pitchRx"), y=0))
```

As you can see, `pitchRx` seems to be slightly below average in terms of popularity. Please help
motivate me to keep the project going by posting an [issue on
github](https://github.com/cpsievert/pitchRx/issues?state=open)!
