---
layout: post
title: "Hello Jekyll"
author: []
categories: [test]
tags: []
reviewer: []
---
{% include JB/setup %}

After some research and contemplation, I've decided to abandon my Wordpress site for this
Jekyll-Bootstrap blog. The main motivation is local file management and producing reproducible
examples of my work to share with others. Here are a few links that motivated and helped complete
the switch:

http://vis.supstat.com/

http://jfisher-usgs.github.io/

http://ramnathv.github.io/poirotBlog/posts/polychart.html

For my own sanity, this post will serve as a reminder of how to post to my own blog:

1. Navigate to the root directory in the terminal and run the following: `$ rake post title="A
Title"`

2. Open the file `_source/yyyy-mm-dd-title.Rmd`, add content and save it!

3. To compile into html: `$ ./_bin/knit _source/yyyy-mm-dd-title.Rmd`

<!-- 4. Copy figures to public Dropbox folder: `$ cp -r
~/Desktop/github/local/blog/_site/figures/yyyy-mm-dd-title ~/Dropbox/Public/blog` -->

5. Preview in all its glory! `$ rake preview`

The output will consist of pretty stuff like this:

```{r test1}
data(mtcars)
summary(mtcars)
```

```{r test2, fig.width=11}
library(ggplot2)
qplot(mpg, wt, color=factor(cyl), data=mtcars)+facet_grid(.~gear, margins=TRUE)
```
