---
layout: post
title: "Visualizing ggplot2 internals with shiny and D3"
author: Carson Sievert
categories: [interactive visualization]
tags: [animint, ggplot2, shiny, D3]
---
{% include JB/setup %}


TL;DR -- I built [this shiny app](https://cpsievert.shinyapps.io/ggtree/) to visualize ggplot2
internals.

I'm fortunate enough to be participating in the Google's Summer of Code program where I'm helping
develop the R package [animint](https://github.com/tdhock/animint). This package is one of
[many](http://ggvis.rstudio.com/) [attempts](http://rcharts.io/) to bring interactive web graphics
to the R console. Animint's approach is somewhat unique in it's translation of ggplot2 code to
HTML/SVG output. To do this, animint first
[compiles](https://github.com/tdhock/animint/wiki/Compiler-details) a list of ggplot objects and
extracts the parts necessary for
[rendering](https://github.com/tdhock/animint/wiki/Renderer-details) output. Although the language
is incredibly expressive and powerful for users, as any ggplot2 developer could tell you, the
structure underlying a ggplot object is quite complicated (sometimes, [even the original author
needs help](https://twitter.com/hadleywickham/status/317279035937923072)).

```{r ggplot2}
library(ggplot2)
p <- ggplot(mtcars, aes(mpg, wt)) +
 geom_point(colour='grey50', size = 4) +
 geom_point(aes(colour = cyl)) + facet_wrap(~am, nrow = 2)
str(p)
```

Of course, this structure gets much more complicated after we "build" the plot --

```{r built}
str(ggplot_build(p))
```

As I started this project, I became frustrated trying to understand/navigate through the nested
list-like structure of ggplot objects. As you can imagine, it isn't an optimal approach to print
out the structure everytime you want to checkout a particular element. Out of this frustration came
an idea to build [this tool](https://cpsievert.shinyapps.io/ggtree/) to help interact with and
visualize this structure. Thankfully, my wonderful GSoC mentor [Toby Dylan
Hocking](http://sugiyama-www.cs.titech.ac.jp/~toby/) agreed that this project could bring value to
the ggplot2 community and encouraged me to pursue it.

By default, this tool presents a [radial Reingold–Tilford
Tree](http://bl.ocks.org/mbostock/4339184) of this nested list structure, but also has options to
use the [collapsable](http://bl.ocks.org/mbostock/4339083) or
[cartesian](http://bl.ocks.org/mbostock/4339184) versions. It also leverages the
[shinyAce](https://github.com/trestletech/shinyAce) package which allows users to send arbitrary
ggplot2 code to a [shiny](http://shiny.rstudio.com/) server thats evaluate the results and
re-renders the visuals. I'm quite happy with the results as I think this tool is a great way to
quickly grasp the internal building blocks of ggplot(s). Please share your thoughts below!
