---
layout: post
title: "Hello Jekyll"
author: [cpsievert]
categories: [test]
---
{% include JB/setup %}

After some research and contemplation, I've decided to abandon my Wordpress site for this Jekyll-Bootstrap blog. The main motivation is local file management and producing reproducible examples of my work to share with others. Here are a few links that motivated and helped complete the switch:

* [http://vis.supstat.com](http://vis.supstat.com)

* [http://jfisher-usgs.github.io](http://jfisher-usgs.github.io)

* [http://ramnathv.github.io/poirotBlog/posts/polychart.html](http://ramnathv.github.io/poirotBlog/posts/polychart.html)

For my own sanity, this post will serve as a reminder of how to post to my own blog:

1. Navigate to the root directory in the terminal and run the following: `$ rake post title="A
Title"`

2. Open the file `_source/yyyy-mm-dd-title.Rmd`, add content and save it!

3. To compile into html: `$ library(knitPost); knitPost("yyyy-mm-dd-title.Rmd")`. Then, if animations are present, add the 'http://cpsievert.github.com/' prefix to src attribute in the markdown file.

4. Preview in all its glory! `$ rake preview`

The output will consist of pretty stuff like this:

```{r test1}
data(mtcars)
summary(mtcars)
```

```{r test2, fig.width=11}
library(ggplot2)
qplot(mpg, wt, color=factor(cyl), data=mtcars)+facet_grid(.~gear, margins=TRUE)
```
