---
title: "Interactive animations of PITCHf/x with animint and pitchRx"
author: "Carson Sievert"
date: "August 26, 2014"
output: html_document
---

**NOTE:** Ironically, some plots below may not render by default if you use Google Chrome, so please use Firefox or Safari. If you insist on Chrome, please close all Chrome instances and then restart Chrome using the flag –allow-file-access-from-files. In Windows, you do this using cmd, the command prompt.

This post demonstrates some new features of the R package [animint](https://github.com/tdhock/animint) (facets, fixed cartesian coordinates, & knitr integration) using [PITCHf/x data](http://en.wikipedia.org/wiki/PITCHf/x). **animint** translates [ggplot2](http://cran.r-project.org/web/packages/ggplot2/index.html)'s grammar of graphics implementation to [SVG](http://en.wikipedia.org/wiki/Scalable_Vector_Graphics) and adds aesthetics/options to enable interactive animations. 

```{r setup, echo = FALSE, message = FALSE}
library(devtools)
library(animint)
library(knitr)
opts_chunk$set(message = FALSE)
```

## Collect and manipulate PITCHf/x data

The R package [pitchRx](https://github.com/cpsievert/pitchRx) provides some utilities for collecting and manipulating PITCHf/x data. **pitchRx** comes prepackaged with every four-seam and cutting fastball from Mariano Rivera and Phil Hughes over the 2011 season (but you can collect whatever you want with `pitchRx::scrape`). **pitchRx** also provides a function `getSnapshots` that returns the spatial location of each pitch over a sequence of time points (in a 3D array). In [this gist](https://gist.github.com/cpsievert/da555f08f3c9ba2c0b8e), I provide a function -- `getLocations` -- that will `getSnapshots`, but return a `data.frame` (which is more convenient for plotting). t can also match an arbitrary number of 'co-variates' with those snapshots, and (optionally) average PITCHf/x parameters over those co-variates.

```{r getLocations}
# This gist requires dplyr, tidyr, & pitchRx; and will install any that are not installed already
source_gist(id = "da555f08f3c9ba2c0b8e")
data(pitches, package = "pitchRx")
dat <- getLocations(pitches, pitcher_name, pitch_type, summarise = TRUE)
head(dat)
```

The `frame` co-variate keeps track of time since the pitcher released the ball (in thousands of a second). Notice that as `frame` becomes larger, the distance from home plate (`y`) becomes smaller.

## Simple animations

By plotting height from the ground (`z`) and the horizontal location relative to home plate (`x`) for each `frame` and stitching those plots into an animation, we acheive the effect of a pitcher throwing the baseball at the viewer. We *could* acheive this by subsetting `dat` and printing a scatterplot of the x/z coordinates for each value of `frame` (which is effectively what `pitchRx::animateFX` does). Or, we can leverage **animint**'s `showSelected` aesthetic and map it to the `frame` variable. This essentially says: 'only show the data corresponding to the currently selected value of `frame`'.

```{r avg_by_plot}
p <- ggplot() + 
  geom_point(aes(x = x, y = z, color = pitch_type, 
                  showSelected = frame), data = dat) + 
  facet_grid(. ~ pitcher_name) + ylim(0, 7) + xlim(-3, 3) +
  coord_equal() + xlab("") + ylab("Height from ground") 
```

If we print this plot, it will actually print a "snapshot" for each value of `frame`.

```{r print_p}
p
```

To convert this plot into an SVG-based animation, we'll group a few animint specific options and the plot itself into a list. 

```{r add_options}
              # defines the plot to convert & specifies its width (in pixels)
plist <- list(strikezone = p + theme_animint(width = 800),            
              # 'animate' over the frame variable
              time = list(variable = "frame", ms = 100),
              # use smooth transitions
              duration = list(frame = 250))
```

If you're using an interactive environment, you can now enter `animint2dir(plist)` or `animint2gist(plist)` to view/share this plot. However, when using [knitr](http://cran.r-project.org/web/packages/knitr/index.html) and R Markdown to create an HTML page (such as this one), just attach a class of 'animint' to your list and **knitr** will know how to embed the plot.

```{r animint}
structure(plist, class = "animint")
```

## Interactive plots

By combining the concept of `showSelected` with a `clickSelects` aesthetic, one can create interaction plots with components that reflect the popular visualization mantra: "[Overview first, zoom and filter, then details on demand](http://www.mat.ucsb.edu/~g.legrady/academic/courses/11w259/schneiderman.pdf)". 

### Overview with zoom & filter

To demonstrate the idea of an overview with a zoom & filter, let's consider a static plot of locations of pitches when they cross home plate (aka a strike-zone plot) and use the `showSelected` aesthetic to filter by the date.

```{r ggstrike}
# reusable function with outline for a faceted strike-zone plot
ggstrike <- function() {
  ggplot() + facet_grid(pitcher_name ~ stand) +
  ylim(0, 7) + xlim(-3, 3) + coord_equal() + xlab("") + 
  ylab("")
}

# overview (not filtered by anything)
strike <- ggstrike() + theme(legend.position = "none") +
          geom_point(data = pitches,
                    aes(x = px, y = pz, color = pitch_type))

# filter by date
strike_date <- ggstrike() + geom_point(data = pitches,
                        aes(x = px, y = pz, color = pitch_type, 
                            showSelected = date))
```

Another plot is required to select dates via a mouseclick. Here we create a time series of the number of pitches thrown (by pitch_type and pitcher_name) and add transparent vertical bars that one may click on to select different dates:

```{r zoom-filter}
# track the number of pitches per game (for each pitcher-type combination)
n_pitches <- pitches %>%
  group_by(pitch_type, pitcher_name, date) %>%
  summarise(count = n()) %>% 
  mutate(group = paste(pitcher_name, pitch_type, sep = ": ")) %>%
  mutate(dated = as.Date(date, format = "%Y_%m_%d")) %>%
  data.frame %>% mutate(max_n = max(count))

# time series plot with clickSelects
series <- ggplot() + 
  geom_line(aes(x = dated, y = count, colour = pitch_type, 
                linetype = pitcher_name, group = group), data = n_pitches) + 
  stat_identity(aes(x = dated, y = max_n, clickSelects = date, alpha = 0.2), 
                geom = "bar", data = n_pitches) + scale_alpha(guide = 'none') +
  xlab("") + ylab("Number of pitches") + theme_animint(width = 800, height = 200)

plist2 <- list(strike = strike,
               strikeDate = strike_date,
               series = series)

structure(plist2, class = "animint")
```

## Interactive animation

Let's combine ideas from the previous two visualizations to create an interactive animination. In the very first visualization, we "summarized" the pitch trajectories over each unique combination of pitcher and pitch type. Now, we take a more granular approach and use filtering to examine the flight of individual pitches. We also add some "details on demand" in the form of a text label giving us the speed of the currently selected pitch.

```{r int-ani}
locs2 <- getLocations(pitches, pitcher_name, pitch_type, date, end_speed, summarise = FALSE) %>%
  mutate(end_speed = paste0("End Speed: ", end_speed, " MPH"))
# Notice a pitch_id co-variate is included automatically to identify which pitch is which
head(locs2)

pitches$pitch_id <- seq_len(dim(pitches)[1])
strike2 <- ggstrike() + theme(legend.position = "none") +
  geom_point(data = pitches, aes(x = px, y = pz, color = pitch_type, 
              showSelected = date, clickSelects = pitch_id))
  
ani <- ggplot() + 
  geom_point(aes(x = x, y = z, color = pitch_type, 
                  showSelected = pitch_id, showSelected2 = frame), 
             data = locs2) + 
  geom_text(aes(x = 0, y = 7, label = end_speed, 
                showSelected = pitch_id, showSelected2 = frame),
            data = locs2) + 
  ylim(0, 7) + xlim(-3, 3) + coord_equal() + xlab("") + 
  ylab("Height from ground")

plist3 <- list(strike2 = strike2,
               ani = ani,
               series = series,
               time = list(variable = "frame", ms = 100),
               duration = list(frame = 250))

structure(plist3, class = "animint")
```

